### 2020-05-11 paper/paper.tex
started the paper

\documentclass{sigchi}

% pandoc setup %

\usepackage[backend=biber]{biblatex}
\addbibresource{references.bib}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% end pandoc setup %

% Use this section to set the ACM copyright statement (e.g. for
% preprints).  Consult the conference website for the camera-ready
% copyright statement.

% Copyright
\CopyrightYear{2020}
%\setcopyright{acmcopyright}
\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}
% DOI
\doi{https://doi.org/10.1145/3313831.XXXXXXX}
% ISBN
\isbn{978-1-4503-6708-0/20/04}
%Conference
\conferenceinfo{CHI'20,}{April  25--30, 2020, Honolulu, HI, USA}
%Price
\acmPrice{\$15.00}

% Use this command to override the default ACM copyright statement
% (e.g. for preprints).  Consult the conference website for the
% camera-ready copyright statement.

%% HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP --
%% Please note you need to make sure the copy for your specific
%% license is used here!
% \toappear{
% Permission to make digital or hard copies of all or part of this work
% for personal or classroom use is granted without fee provided that
% copies are not made or distributed for profit or commercial advantage
% and that copies bear this notice and the full citation on the first
% page. Copyrights for components of this work owned by others than ACM
% must be honored. Abstracting with credit is permitted. To copy
% otherwise, or republish, to post on servers or to redistribute to
% lists, requires prior specific permission and/or a fee. Request
% permissions from \href{mailto:Permissions@acm.org}{Permissions@acm.org}. \\
% \emph{CHI '16},  May 07--12, 2016, San Jose, CA, USA \\
% ACM xxx-x-xxxx-xxxx-x/xx/xx\ldots \$15.00 \\
% DOI: \url{http://dx.doi.org/xx.xxxx/xxxxxxx.xxxxxxx}
% }

% Arabic page numbers for submission.  Remove this line to eliminate
% page numbers for the camera ready copy
% \pagenumbering{arabic}

% Load basic packages
\usepackage{balance}       % to better equalize the last page
\usepackage{graphics}      % for EPS, load graphicx instead
\usepackage[T1]{fontenc}   % for umlauts and other diaeresis
\usepackage{txfonts}
\usepackage{mathptmx}
\usepackage[pdflang={en-US},pdftex]{hyperref}
\usepackage{color}
\usepackage{booktabs}
\usepackage{textcomp}

% Some optional stuff you might like/need.
\usepackage{microtype}        % Improved Tracking and Kerning
% \usepackage[all]{hypcap}    % Fixes bug in hyperref caption linking
\usepackage{ccicons}          % Cite your images correctly!
% \usepackage[utf8]{inputenc} % for a UTF8 editor only

% If you want to use todo notes, marginpars etc. during creation of
% your draft document, you have to enable the "chi_draft" option for
% the document class. To do this, change the very first line to:
% "\documentclass[chi_draft]{sigchi}". You can then place todo notes
% by using the "\todo{...}"  command. Make sure to disable the draft
% option again before submitting your final document.
\usepackage{todonotes}

% Paper metadata (use plain text, for PDF inclusion and later
% re-using, if desired).  Use \emtpyauthor when submitting for review
% so you remain anonymous.
\def\plaintitle{Abstract Program Visualization for Model-View-Update
User Interfaces}
\def\plainauthor{Geoffrey Litt}
\def\emptyauthor{}
\def\plainkeywords{Program visualization, program understanding, debugging}
\def\plaingeneralterms{Program visualization, program understanding, debugging}

% llt: Define a global style for URLs, rather that the default one
\makeatletter
\def\url@leostyle{%
  \@ifundefined{selectfont}{
    \def\UrlFont{\sf}
  }{
    \def\UrlFont{\small\bf\ttfamily}
  }}
\makeatother
\urlstyle{leo}

% To make various LaTeX processors do the right thing with page size.
\def\pprw{8.5in}
\def\pprh{11in}
\special{papersize=\pprw,\pprh}
\setlength{\paperwidth}{\pprw}
\setlength{\paperheight}{\pprh}
\setlength{\pdfpagewidth}{\pprw}
\setlength{\pdfpageheight}{\pprh}

% Make sure hyperref comes last of your loaded packages, to give it a
% fighting chance of not being over-written, since its job is to
% redefine many LaTeX commands.
\definecolor{linkColor}{RGB}{6,125,233}
\hypersetup{%
  pdftitle={\plaintitle},
% Use \plainauthor for final version.
%  pdfauthor={\plainauthor},
  pdfauthor={\plainauthor},
  pdfkeywords={\plainkeywords},
  pdfdisplaydoctitle=true, % For Accessibility
  bookmarksnumbered,
  pdfstartview={FitH},
  colorlinks,
  citecolor=black,
  filecolor=black,
  linkcolor=black,
  urlcolor=linkColor,
  breaklinks=true,
  hypertexnames=false
}

% create a shortcut to typeset table headings
% \newcommand\tabhead[1]{\small\textbf{#1}}

% End of preamble. Here it comes the document.
\begin{document}

\title{\plaintitle}

\numberofauthors{1}
\author{%
  \alignauthor{Geoffrey Litt}\\
    \affaddr{MIT CSAIL}\\
    \email{glitt@mit.edu}\\
}

\maketitle

\begin{abstract}
  This is the abstract
\end{abstract}

% ACM Classfication

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003120.10003121</concept_id>
<concept_desc>Human-centered computing~Human computer interaction (HCI)</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003120.10003121.10003125.10011752</concept_id>
<concept_desc>Human-centered computing~Haptic devices</concept_desc>
<concept_significance>300</concept_significance>
</concept>
<concept>
<concept_id>10003120.10003121.10003122.10003334</concept_id>
<concept_desc>Human-centered computing~User studies</concept_desc>
<concept_significance>100</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Human-centered computing~Human computer interaction (HCI)}
\ccsdesc[300]{Human-centered computing~Haptic devices}
\ccsdesc[100]{Human-centered computing~User studies}

% Author Keywords
\keywords{\plainkeywords}

% Print the classficiation codes
\printccsdesc
Please use the 2012 Classifiers and see this link to embed them in the text: \url{https://dl.acm.org/ccs/ccs_flat.cfm}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Lots of recent program vis work is low level. Tied directly to the
source code, visualizing state at individual lines, individual
variables. Makes sense for novices: small programs, need help
understanding small details.

OTOH: many tasks require higher level of understanding system behavior.
Eg, initial explanation of a codebase. How do we make a ``whiteboard
drawing'' live? Some challenges:

\begin{itemize}
\tightlist
\item
  relevant state isn't necessarily individual variable values: need to
  select certain relevant attributes to visualize. Corollary: can't be
  as automated as low level visualization.
\item
  Too much work on ``targeted debugging'', not enough on generalized
  understanding of a system. More specifically: if we don't know what
  change you might want to make to a codebase, how can we maximally
  equip you to be ready to make an arbitrary change?

  \begin{itemize}
  \tightlist
  \item
    Naur Programming as theory building
  \end{itemize}
\item
  combines both static + dynamic aspects: code modules, abstracted state
\item
  Lots of non-numeric data
\end{itemize}

Abstract visualization is an interesting direction. Can be
super-specific, eg algorithm animation \autocite{brown1984,stasko1990}
or class and thread vis of \autocite{reiss2003,reiss2005}.

But a major challenge is making it easy enough for programmer to
actually make it worth it \autocite{reiss2007}

This is an initial exploration. Some solution characteristics:

\begin{itemize}
\tightlist
\item
  redux apps \autocite{czaplicki,fowler2020}: 1) broader than Vega,
  narrower than general programs, 2) imposes a worldview of abstract
  state and simplified event stream. THIS IS THE HEART OF THE WORK

  \begin{itemize}
  \tightlist
  \item
    In this work: focus on TodoMVC specifically, but in theory should
    generalize to Redux programs. (Future work: how to efficiently
    generate a visualization like this one)
  \end{itemize}
\item
  ``Guided tour'': introducing you to how the application works.
\end{itemize}

\hypertarget{sec:related-work}{%
\section{Related Work}\label{sec:related-work}}

\begin{itemize}
\tightlist
\item
  low level program vis

  \begin{itemize}
  \tightlist
  \item
    Learnable \autocite{victora}
  \item
    omnicode \autocite{kang2017}
  \item
    Projection boxes \autocite{lerner2020}
  \item
    Theia \autocite{pollock2019}
  \item
    Theseus \autocite{lieber2014}
  \end{itemize}
\item
  Whyline, targeted interrogation \autocite{ko2004}
\item
  In-situ: nice taxonomy of visualizations (but still limited to Vega,
  narrow domain) \autocite{hoffswell2018a}
\item
  Myers taxonomy \autocite{myers1990}
\item
  Redux / Elm

  \begin{itemize}
  \tightlist
  \item
    redux dev tools Tree viewer
  \end{itemize}
\item
  http://cs.brown.edu/\textasciitilde spr/research/bloom/jvlexec.pdf
\item
  Steve Reiss overview
\item
  Brad Myers incense
\item
  Girba's Mondrian: a toolkit for programmers building vis
  \autocite{meyer2006}
\end{itemize}

\hypertarget{methods}{%
\subsection{Methods}\label{methods}}

Only some initial experiments.

Redux dev tools monitor D3 + react

Show the mockups Developed 3-4 sparkline style visualizations

\begin{itemize}
\tightlist
\item
  line graph
\item
  Categorical graph
\item
  Collection dots view?
\item
  String change view?
\end{itemize}

\hypertarget{results}{%
\subsection{Results}\label{results}}

\begin{itemize}
\tightlist
\item
  get feedback from 1-2 people?
\item
  Tried using it myself
\end{itemize}

\hypertarget{discussion}{%
\subsection{Discussion}\label{discussion}}

\hypertarget{future-work}{%
\subsection*{Future work}\label{future-work}}
\addcontentsline{toc}{subsection}{Future work}

% Balancing columns in a ref list is a bit of a pain because you
% either use a hack like flushend or balance, or manually insert
% a column break.  http://www.tex.ac.uk/cgi-bin/texfaq2html?label=balance
% multicols doesn't work because we're already in two-column mode,
% and flushend isn't awesome, so I choose balance.  See this
% for more info: http://cs.brown.edu/system/software/latex/doc/balance.pdf
%
% Note that in a perfect world balance wants to be in the first
% column of the last page.
%
% If balance doesn't work for you, you can remove that and
% hard-code a column break into the bbl file right before you
% submit:
%
% http://stackoverflow.com/questions/2149854/how-to-manually-equalize-columns-
% in-an-ieee-paper-if-using-bibtex
%
% Or, just remove \balance and give up on balancing the last page.
%
\balance{}

% \bibliographystyle{SIGCHI-Reference-Format}
% \bibliography{sample}

\printbibliography

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2020-05-11 paper/compile-html.sh

@@ -1,0 +1,16 @@
+set -e
+
+ pandoc \
+  --filter conditional-render \
+  --filter pandoc-crossref \
+  --filter pandoc-citeproc \
+  --metadata=format:html \
+  -s \
+  --number-sections \
+  -o paper.html \
+  --css basic.css \
+  --toc \
+  --toc-depth=1 \
+  --variable=toc-title:"Contents" \
+  --template=pandoc-template-html.html \
+  paper.md

### 2020-05-11 paper/compile-latex.sh

@@ -1,0 +1,19 @@
+set -e
+
+pandoc \
+  --filter conditional-render \
+  --filter pandoc-crossref \
+  --filter pandoc-citeproc \
+  -s \
+  -o \
+  paper.tex \
+  --template=chi.tex \
+  --biblatex \
+  --metadata=format:pdf \
+  paper.md
+
+pdflatex paper.tex
+biber paper
+pdflatex -interaction=batchmode paper.tex
+pdflatex -interaction=batchmode paper.tex
+open paper.pdf

### 2020-05-11 paper/watch.sh

@@ -1,0 +1,10 @@
+set -e
+
+# kill any existing servers
+ps aux | grep node | grep browser-sync | awk '{ print $2}' | xargs kill
+
+ls *.md | entr -s "./compile-html.sh && ./compile-latex.sh" &
+browser-sync start --server --files paper.html --no-notify --no-open --port 9000 &
+
+open "http://localhost:9000/paper.html"
+

### 2020-05-12 src/containers/DevTools.js
tweaking

@@ -21,6 +21,7 @@ const DevTools = createDevTools(
     changeMonitorKey='ctrl-m'
     defaultIsVisible={true}
     defaultSize={0.3}
+    defaultPosition="bottom"
   >
     <VisMonitor />
     <LogMonitor theme="tomorrow" />

### 2020-05-13 paper/paper.tex
update paper
tweaking
more work on paper

@@
 \begin{abstract}
-  This is the abstract
+  Visualizing the runtime behavior of programs can help programmers with
+  targeted debugging and general understanding. For understanding
+  complex programs, visualizations abstracted from the low-level code
+  are most helpful, but this introduces new challenges: how does the
+  programmer specify what to visualize, and how do we visualize complex
+  data structures which aren't just primitive values? In this work, I
+  present an approach to visualizing the behavior of user interfaces
+  built with the Model-View-Update pattern. I present a prototype
+  runtime visualization system built on the Redux library and argue
+  that, by exploiting the natural abstraction characteristics of this
+  application architecture, we can create useful runtime visualizations
+  with minimal programmer effort.
 \end{abstract}
@@
 % Print the classficiation codes
-\printccsdesc
-Please use the 2012 Classifiers and see this link to embed them in the text: \url{https://dl.acm.org/ccs/ccs_flat.cfm}
+% \printccsdesc
+% Please use the 2012 Classifiers and see this link to embed them in the text: \url{https://dl.acm.org/ccs/ccs_flat.cfm}
 
@@
 
-Lots of recent program vis work is low level. Tied directly to the
-source code, visualizing state at individual lines, individual
-variables. Makes sense for novices: small programs, need help
-understanding small details.
+Much recent work in program visualization
+\autocite{victora,guo2013,hoffswell2018a,pollock2019,kasibatla2018}
+focuses on low-level details: showing the values of individual
+variables, connected to individual lines of source code. This makes
+sense for small programs, and for helping novices understand the basics
+of programming languages. But these visualizations don't address the
+needs of more experienced programmers working with larger programs.
+Gaining a general understanding of a large program requires zooming out
+far above the source code level.
 
-OTOH: many tasks require higher level of understanding system behavior.
-Eg, initial explanation of a codebase. How do we make a ``whiteboard
-drawing'' live? Some challenges:
+This leads to the idea of \emph{abstract program visualization}:
+creating abstract, program-specific views of runtime state or static
+code, to help someone understand the program. This idea has been
+explored in the context of teaching how algorithms work
+\autocite{brown1984,stasko1990} and in understanding the behavior of
+multithreaded Java programs \autocite{reiss2003,reiss2005}. In this
+context, a tricky challenge emerges \autocite{reiss2007}: how can we
+simultaneously tailor the visualization to the specific domain, while
+enabling the programmer to create the visualization with minimal effort?
+On the one hand, overly generic visualizations (as used in low-level
+visualization systems like Python Tutor \autocite{guo2013}) will often
+fail to capture the higher-level context of the specific program. On the
+other hand, if a visualization is too specific and takes too much work
+to create, it won't be reasonably possible for programmers to create the
+visualization.
+
+I think a promising strategy for approaching this problem is to create
+runtime visualization systems coupled to a particular domain-specific
+framework or language. Frameworks and DSLs often impose a particular
+mental model, code architecture style, and other constraints that
+usefully narrow the space of possible programs relative to a
+general-purpose language. On the other hand, there are still many
+different programs that can be built in one framework, so the effort of
+building a visualization system can be amortized over thousands of
+programs rather than concentrated on a single one.
+
+One recent example of this strategy is visualizing runtime state in Vega
+\autocite{hoffswell2018a}, a data visualization language. Because the
+language is so domain-specific, the runtime visualization system is able
+to make many contextual assumptions leading to a useful visualization.
+
+In this work, I propose a runtime program visualization system for a
+slightly more general category of applications: user interfaces built
+with the Model-View-Update (MVU) architecture \autocite{fowler2020},
+also commonly known as the Elm Architecture \autocite{czaplicki}. MVU
+encourages the state of the interface to be centralized in a single data
+structure, derived by running a pure reducer function over a stream of
+events.
+
+This architecture has been found to have many practical benefits for
+program understanding and developer experience (e.g., making
+``time-travel debugging'' relatively trivial to implement), but I think
+it also has useful characteristics for abstract program visualization as
+well. In particular, MVU naturally encourages programmers to define
+abstractions that represent the essence of their application: 1) a
+stream of semantically meaningful events, representing important changes
+in the application, 2) a state object that represents all the core state
+of the UI, organized in a hierarchical way. My hypothesis is that it's
+possible to build a system that can visualize MVU interfaces with
+relatively little additional effort from the programmer, because of
+these natural characteristics of the architecture.
+
+To begin to explore this idea, I've prototyped a runtime visualization
+system on top of the popular Redux \autocite{zotero-621} Javascript
+library. Within the limited scope of this project, I've focused
+specifically on making a prototype specifically designed to visualize
+the state of the TodoMVC demo application. I've designed some
+visualization formats tailored to the state of that application, and
+through my own usage I've begun to gain a preliminary understanding what
+kinds of visualizations might be useful to programmers navigating the
+state of MVU applications over time.
+
+Much future work remains to fully flesh out this idea, including:
 
@@
 \item
-  relevant state isn't necessarily individual variable values: need to
-  select certain relevant attributes to visualize. Corollary: can't be
-  as automated as low level visualization.
+  developing a crisper understanding of the needs of programmers who are
+  debugging MVU applications or trying to gain a general understanding
+  of their behavior
 \item
-  Too much work on ``targeted debugging'', not enough on generalized
-  understanding of a system. More specifically: if we don't know what
-  change you might want to make to a codebase, how can we maximally
-  equip you to be ready to make an arbitrary change?
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Naur Programming as theory building
-  \end{itemize}
-\item
-  combines both static + dynamic aspects: code modules, abstracted state
-\item
-  Lots of non-numeric data
-\end{itemize}
-
-Abstract visualization is an interesting direction. Can be
-super-specific, eg algorithm animation \autocite{brown1984,stasko1990}
-or class and thread vis of \autocite{reiss2003,reiss2005}.
-
-But a major challenge is making it easy enough for programmer to
-actually make it worth it \autocite{reiss2007}
-
-This is an initial exploration. Some solution characteristics:
-
-\begin{itemize}
-\tightlist
-\item
-  redux apps \autocite{czaplicki,fowler2020}: 1) broader than Vega,
-  narrower than general programs, 2) imposes a worldview of abstract
-  state and simplified event stream. THIS IS THE HEART OF THE WORK
-
-  \begin{itemize}
-  \tightlist
-  \item
-    In this work: focus on TodoMVC specifically, but in theory should
-    generalize to Redux programs. (Future work: how to efficiently
-    generate a visualization like this one)
-  \end{itemize}
-\item
-  ``Guided tour'': introducing you to how the application works.
+  generalizing the system so that it actually works with many Redux
+  applications instead of just one, and has a way for a programmer to
+  create a program-specific visualization
 \end{itemize}
@@
 
-\begin{itemize}
-\tightlist
-\item
-  low level program vis
+Reiss \autocite{reiss2007} provides a useful taxonomy of execution
+visualizations, with pointers to prior research. Some particularly
+relevant dimensions for this work include abstract vs concrete, and
+effort required to create the visualization.
+
+Many systems have explored visualizing execution state at the level of
+individual source lines, including Learnable Programming
+\autocite{victora}, Python Tutor \autocite{guo2013}, Omnicode
+\autocite{kang2017}, and Theia \autocite{pollock2019}.
+
+Some systems have explored somewhat more abstract views. Projection
+Boxes {[}\textcite{lerner2020} provides a way of selectively showing
+parts of application state, and Seymour \autocite{kasibatla2018}
+provides a ``macro'' visualization to generally show the layout of
+execution flow, in addition to a ``micro'' visualization.
+
+Other systems have explored fully abstract program visualization,
+entirely disconnected from the source code. For example, Balsa
+\autocite{brown1984} and Tango \autocite{stasko1990} show animated views
+of algorithms operating, and Jive \autocite{reiss2003} and Jove
+\autocite{reiss2005} visualize various high-level projections of the
+execution of Java programs, e.g.~when different threads are running.
+
+Hoffswell et al propose a system for visualizing runtime state inside
+Vega data visualizations \autocite{hoffswell2018a}. That work fits into
+the category of visualizing state next to source code lines, but by
+integrating with a very high level domain-specific language, achieves
+more abstraction than visualization systems for general languages like
+Python. They also propose a design space for visualizations embedded in
+source code, which I plan to build on in this work.
+
+\hypertarget{sec:design}{%
+\section{Visualization design}\label{sec:design}}
+
+\hypertarget{use-cases}{%
+\subsection{Use cases}\label{use-cases}}
+
+I had some prior experience with the Redux Dev Tools system, which
+provides the ability to inspect application state in Redux applications.
+From this personal experience I identified two distinct use cases for a
+runtime visualization:
 
@@
 \item
-    Learnable \autocite{victora}
+  \textbf{Localizing within a trace}: \emph{Where do I need to rewind
+  to, in order to inspect a particularly relevant point in an execution
+  trace?} This is most often helpful when debugging a particular
+  problem. Scrubbing back and forth while watching the UI change is a
+  manageable solution, but it's inefficient, and sometimes the relevant
+  state isn't directly visible in the UI.
 \item
-    omnicode \autocite{kang2017}
-  \item
-    Projection boxes \autocite{lerner2020}
-  \item
-    Theia \autocite{pollock2019}
-  \item
-    Theseus \autocite{lieber2014}
-  \end{itemize}
-\item
-  Whyline, targeted interrogation \autocite{ko2004}
-\item
-  In-situ: nice taxonomy of visualizations (but still limited to Vega,
-  narrow domain) \autocite{hoffswell2018a}
-\item
-  Myers taxonomy \autocite{myers1990}
-\item
-  Redux / Elm
-
-  \begin{itemize}
-  \tightlist
-  \item
-    redux dev tools Tree viewer
-  \end{itemize}
-\item
-  http://cs.brown.edu/\textasciitilde spr/research/bloom/jvlexec.pdf
-\item
-  Steve Reiss overview
-\item
-  Brad Myers incense
-\item
-  Girba's Mondrian: a toolkit for programmers building vis
-  \autocite{meyer2006}
+  \textbf{Generally understanding program behavior over time}:
+  \emph{Overall, what happened as I interacted with the program?}
+  Sometimes I'm not debugging a particular problem, and I'm more
+  interested in just seeing general information about how a program is
+  behaving over time. This is helpful when explaining the system's
+  behavior to a new programmer who's preparing to work on the system, or
+  for building up my own passive understanding of a system.
 \end{itemize}
 
-\hypertarget{methods}{%
-\subsection{Methods}\label{methods}}
+These two goals partially overlap, but can also lead in different design
+directions. For example, localizing a specific point in a trace can
+benefit from a more active interrogatory approach (e.g.~as explored in
+the Whyline system \autocite{ko2004}), but general program understanding
+might benefit from a more passive interaction style.
 
-Only some initial experiments.
+\hypertarget{data-structures}{%
+\subsection{Data structures}\label{data-structures}}
 
-Redux dev tools monitor D3 + react
+Many concrete and low-level program visualizations focus on showing
+primitive values, especially numeric values. However, the state of an
+arbitrary MVU application often contains complex nested data structures,
+which contain many non-numeric values: booleans, strings, and enum
+values. One challenge for this system is to find ways to visualize these
+types of structures.
 
-Show the mockups Developed 3-4 sparkline style visualizations
+\hypertarget{sec:design}{%
+\section{Visualization Design}\label{sec:design}}
 
-\begin{itemize}
-\tightlist
-\item
-  line graph
-\item
-  Categorical graph
-\item
-  Collection dots view?
-\item
-  String change view?
-\end{itemize}
+\hypertarget{context-todomvc}{%
+\subsection{Context: TodoMVC}\label{context-todomvc}}
 
-\hypertarget{results}{%
-\subsection{Results}\label{results}}
+In order to focus my effort on concretely understanding the utility of
+visualizations, rather than building out infrastructure, I built a
+visualization system for a specific application: the Redux
+implementation of the TodoMVC GUI benchmark. TodoMVC is a basic todo
+list UI where you can add, edit, delete, and complete todos. You can
+filter the list of todos shown to either active or completed ones.
 
-\begin{itemize}
-\tightlist
-\item
-  get feedback from 1-2 people?
-\item
-  Tried using it myself
-\end{itemize}
+In Redux, the implementation stores an app state object as shown in
+Figure~\ref{fig:todomvc-state}. There are Redux actions corresponding to
+each of the main actions listed above, e.g.~``add todo'' and ``set
+visibility filter''. Importantly, the Redux events capture an abstract,
+semantically meaningful picture of the user's interactions: when adding
+a new todo, the user's keystrokes are collected in the local state of a
+React component, and only a single ``add todo'' event is triggered in
+Redux once the user finally adds the new todo.
 
-\hypertarget{discussion}{%
-\subsection{Discussion}\label{discussion}}
+\begin{figure}
+\hypertarget{fig:todomvc-state}{%
+\centering
+\includegraphics[width=1.82292in,height=2.08333in]{images/todomvc-state.png}
+\caption{TodoMVC Redux application state}\label{fig:todomvc-state}
+}
+\end{figure}
 
-\hypertarget{future-work}{%
-\subsection*{Future work}\label{future-work}}
-\addcontentsline{toc}{subsection}{Future work}
+\hypertarget{overall-layout}{%
+\subsection{Overall layout}\label{overall-layout}}
+
+\begin{figure}
+\hypertarget{fig:mockup}{%
+\centering
+\includegraphics[width=3.4375in,height=2.25in]{images/mockup.png}
+\caption{JSON tree with inline sparklines}\label{fig:mockup}
+}
+\end{figure}
+
+My initial idea, as shown in Figure~\ref{fig:mockup}, was to show the
+current state of the application as a nested JSON tree, and then to show
+small sparkline-style visualizations next to nodes of the tree. This
+design draws some inspiration from \autocite{hoffswell2018a}, but
+differs in that it uses visualization to annotate the application's
+state tree, rather than its source code. The advantage of this design is
+that it closely and directly links the current state to data from the
+execution history, but that link also causes thorny problems---for
+example, how do you deal with nodes that have disappeared from the
+current state? Perhaps more concerningly, by tying the visualization to
+the \emph{concrete} current state, it limits the ability of the
+programmer to create a customized abstract view. Linking the
+visualization to the Redux state is a much more abstract view than
+linking to lines of source code, but it still presents a limitation.
+
+As a result, in my next iteration I switched to a different layout,
+shown in Figure~\ref{fig:timeline}: a vertically stacked list of small
+visualizations of state over time. Each visualization can display an
+arbitrary projection of the app's Redux state. Because the graphs are
+horizontally aligned, it's easy to see how different aspects of the
+app's state have changed in relation to each other. While I haven't
+implemented this yet, I imagine that programmers would be able to
+dynamically add visualizations to this list, specifying useful
+projections of app state, and deciding what type of visualization to use
+for each projection.
+
+\begin{figure}
+\hypertarget{fig:timeline}{%
+\centering
+\includegraphics[width=3.4375in,height=2.08333in]{images/timeline.png}
+\caption{Timeline view of stacked visualizations}\label{fig:timeline}
+}
+\end{figure}
+
+One thing lost in the timeline view is the concrete view of the app's
+entire state. It's still useful to see this, so I added a separate panel
+which displays that data. The user can scrub through history in the
+timeline, ``pin'' the app at a particular point in time, and then use
+the separate state view to drill into the app's concrete state at that
+point.
+
+\hypertarget{visualization-types}{%
+\subsection{Visualization Types}\label{visualization-types}}
+
+Here are the specific types of visualizations I prototyped (all shown in
+Figure~\ref{fig:timeline}):
+
+\emph{Action list}: I found that skimming a list of actions represented
+as text (``ADD\_TODO'', ``EDIT\_TODO'') required a lot of conscious
+reading effort. Instead, by choosing a colorful symbol for each action
+in the app, we can take advantage of pre-attentive processing to more
+quickly understand what actions have occurred in the execution trace. In
+this case I chose symbols for all the actions in TodoMVC; more
+generally, a programmer could specify a meaningful symbol for every
+action in their application. In some cases it might be difficult to
+choose meaningful and different symbols for all actions; falling back to
+random symbols or colored dots could work as well. In using this tool
+I've found that the symbolic action list makes it far easier to find a
+point in an execution trace that I'm looking for.
+
+\emph{Line graph}: This is simply a line graph of some numeric quantity
+over time. In this context I've used it to visualize quantities like
+``Number of todos visible''. Choosing a y-axis is quite tricky because
+the full range of values can't be known in advance. In trying out
+different options and using the tool myself, I decided that viewing
+relative changes over time was most important---generally I'm looking
+for things like ``when did the number of todos go down?''. Therefore, I
+let each graph scale to the current range of values and don't even show
+a y-axis label---I'm not aiming to precisely read numeric values off the
+graph.
+
+\emph{Enum graph}: User interfaces commonly have enums / union types,
+which can take on a small number of predefined values. To represent enum
+values changing over time, I chose to use both a color and position
+encoding, as a way of redundantly encoding the information and .
+
+With more time, I'd like to explore many other types of visualizations
+in addition to these. One particular interest is displaying the entire
+state of a collection of objects in a single graph.
+
+\hypertarget{prototype-implementation}{%
+\subsection{Prototype Implementation}\label{prototype-implementation}}
+
+I implemented a working prototype version of this visualization system.
+I built a plugin on top of the existing Redux Dev Tools, which provides
+substantial infrastructure for inspecting and manipulating the state of
+a Redux application.
+
+I used the React and Redux frameworks to implement the main skeleton of
+my system. The graphs are built in a combination of d3 and React---using
+d3 for computing scales and positions, and React for SVG rendering.
+
+\hypertarget{sec:discussion}{%
+\subsection{Discussion and Future Work}\label{sec:discussion}}
+
+This work is still an early prototype and there are many opportunities
+for future work.
+
+Using this system myself, I found that I was able to more quickly get an
+overall sense of what happened in an execution trace by looking at these
+visualizations than looking at the existing Redux Dev Tools display.
+However, I want to gain a clearer understanding of what questions people
+have when learning about the behavior of a UI, in order to evaluate the
+usefulness of the system. In particular, I'm curious about general
+``program understanding'' as opposed to targeted debugging. Could this
+visualization be a useful aid when onboarding someone into a codebase
+and teaching them how it works?
+
+There's lots of future work to refine the core visualizations further. I
+haven't yet explored visualizing a complex object in a single graph, or
+showing strings changing over time. I'd also like to more clearly
+incorporate Hoffswell et al's taxonomy \autocite{hoffswell2018a} into
+this work, evaluating these visualizations in those terms and explicitly
+extending that taxonomy.
+
+Another area of work is generalizing this system to work with any Redux
+application. I'd like to explore the programmer experience of creating
+these visualizations for an existing complex application. How much of
+that process can be automated? How can we make it easy for the
+programmer to decide which visualizations would be helpful, and then to
+actually specify those visualizations? As an initial idea, I imagine
+that the programmer could specify an arbitrary expression over the Redux
+state,choose from a predefined list of visualizations for showing the
+output of that expression, and then add that to the timeline panel in
+this tool.
+
+Program visualization offers a rich set of possibilities for helping
+people understand their code better. In this work, I've provided an
+initial prototype of a system for visualizing the runtime state of
+Model-View-Update user interfaces, exploiting the natural architecture
+of these applications to show an abstract picture of code execution over
+time.
 

### 2020-05-13 precache-manifest.1c9237cf565ee6320148032c0953ccc0.js
Updates

@@ -1,19 +1,19 @@
 self.__precacheManifest = (self.__precacheManifest || []).concat([
   {
-    "revision": "a5b6ff5dc9b4167423f66d35b9b14053",
+    "revision": "e8e9f154ac067d44cd81b38b8311cf09",
     "url": "/pages/6894-sp20/FP-Program-Execution-Visualization/index.html"
   },
   {
-    "revision": "6b24a57707279fbf9cda",
+    "revision": "4f2b25ca732896f8f1fe",
     "url": "/pages/6894-sp20/FP-Program-Execution-Visualization/static/css/2.8dd801da.chunk.css"
   },
   {
-    "revision": "6b24a57707279fbf9cda",
-    "url": "/pages/6894-sp20/FP-Program-Execution-Visualization/static/js/2.5158c329.chunk.js"
+    "revision": "4f2b25ca732896f8f1fe",
+    "url": "/pages/6894-sp20/FP-Program-Execution-Visualization/static/js/2.052aba8e.chunk.js"
   },
   {
-    "revision": "0269df3f8b02e26626c7",
-    "url": "/pages/6894-sp20/FP-Program-Execution-Visualization/static/js/main.085b9e87.chunk.js"
+    "revision": "bdccee12cab5f02dc7d3",
+    "url": "/pages/6894-sp20/FP-Program-Execution-Visualization/static/js/main.47d86621.chunk.js"
   },
   {
     "revision": "1c8fb0e2dade0b210746",

### 2020-05-13 service-worker.js

@@ -14,7 +14,7 @@
 importScripts("https://storage.googleapis.com/workbox-cdn/releases/4.3.0/workbox-sw.js");
 
 importScripts(
-  "/pages/6894-sp20/FP-Program-Execution-Visualization/precache-manifest.dbec105eb0ae008fa8f16e6af7e1b5ab.js"
+  "/pages/6894-sp20/FP-Program-Execution-Visualization/precache-manifest.1c9237cf565ee6320148032c0953ccc0.js"
 );
 
 self.addEventListener('message', (event) => {

### 2020-05-13 static/js/main.085b9e87.chunk.js

@@ -1,2 +1,0 @@
-(window.webpackJsonp=window.webpackJsonp||[]).push([[0],{287:function(e,t,n){e.exports=n(542)},542:function(e,t,n){"use strict";n.r(t);var a={};n.r(a),n.d(a,"addTodo",function(){return me}),n.d(a,"deleteTodo",function(){return pe}),n.d(a,"editTodo",function(){return he}),n.d(a,"completeTodo",function(){return ve}),n.d(a,"completeAllTodos",function(){return ge}),n.d(a,"clearCompleted",function(){return be}),n.d(a,"setVisibilityFilter",function(){return Se});var r=n(0),o=n.n(r),c=n(69),i=n(10),l=n(8),u=n(15),d=n(223),s=n.n(d),f=n(224),m=n.n(f),p=n(225),h=n.n(p),v=n(38),g=n(39),b=n(43),S=n(40),E=n(44),I=n(6),y=n(1),O=n.n(y),j=n(7);function w(){var e=Object(I.a)(["\n  display: inline-block;\n  width: 80px;\n  background-color: ",";\n  overflow: hidden;\n  text-overflow: ellipsis;\n  margin-right: 10px;\n  padding: 5px;\n  border-radius: 5px;\n  cursor: pointer;\n"]);return w=function(){return e},e}var T=j.a.div(w(),function(e){return e.selected?"#ffcece":e.active?"#a6d2ff":"#eee"}),k=function(e){var t=e.currentStateId,n=e.stagedActionIds,a=e.actionsById,r=e.selectedStateId,c=e.setSelectedStateId,i=e.jumpToState,l=n.map(function(e){var n=a[e];return o.a.createElement(T,{key:e,onMouseOver:function(){return c(e)},selected:e===r,active:t>=e,onClick:function(){return i(e)}},n.action.type)});return o.a.createElement("div",null,l)},C=n(76),x=n(2),A=function(e){var t=e.path,n=e.stroke,a=e.fill,r=e.strokeWidth;return o.a.createElement("path",{d:t,fill:a,stroke:n,strokeWidth:r})};A.defaultProps={stroke:"blue",fill:"none",strokeWidth:2};var N=A;function _(){var e=Object(I.a)(["\n  display: inline-block;\n  color: #aaa;\n"]);return _=function(){return e},e}function F(){var e=Object(I.a)(["\n  border: solid thin #eee;\n  cursor: pointer;\n  margin-right: 10px;\n"])
[…]

### 2020-05-13 static/js/main.47d86621.chunk.js

@@ -1,0 +1,2 @@
+(window.webpackJsonp=window.webpackJsonp||[]).push([[0],{287:function(e,t,n){e.exports=n(542)},542:function(e,t,n){"use strict";n.r(t);var a={};n.r(a),n.d(a,"addTodo",function(){return he}),n.d(a,"deleteTodo",function(){return ve}),n.d(a,"editTodo",function(){return ge}),n.d(a,"completeTodo",function(){return Ee}),n.d(a,"completeAllTodos",function(){return Se}),n.d(a,"clearCompleted",function(){return be}),n.d(a,"setVisibilityFilter",function(){return Ie});var r=n(0),o=n.n(r),i=n(71),c=n(9),l=n(7),u=n(15),d=n(223),s=n.n(d),m=n(224),f=n.n(m),p=n(225),h=n.n(p),v=n(38),g=n(39),E=n(44),S=n(40),b=n(45),I=n(10),O=n(1),y=n.n(O),T=n(5),j=n(43),w=n(2),x=function(e){var t=e.path,n=e.stroke,a=e.fill,r=e.strokeWidth,i=e.strokeDasharray;return o.a.createElement("path",{d:t,fill:a,stroke:n,strokeWidth:r,strokeDasharray:i,opacity:.8})};x.defaultProps={stroke:"blue",fill:"none",strokeWidth:2};var k=x;function C(){var e=Object(I.a)(["\n  border: solid thin #eee;\n  cursor: pointer;\n  overflow: visible;\n"]);return C=function(){return e},e}var _=T.a.svg(C()),D=function(e){var t,n,a=e.data,r=e.width,i=e.height,c=e.setSelectedStateId,l=e.selectedStateId,u=e.jumpToState,d=e.currentStateId,s=e.resetToSelectedState,m=a.slice(-1)[0],f=a.concat([{stateId:m.stateId+1,value:m.value}]),p=w.h().domain(w.e(f,function(e){return e.stateId})).range([1,r-1]),h=w.h().domain(w.e(f,function(e){return e.value})).range([i-1,1]),v=w.f().defined(function(e){return!isNaN(e.value)}).curve(w.d).x(function(e){return p(e.stateId)}).y(function(e){return h(e.value)}),g=w.f().curve(w.c)([[p(d),h.range()[0]],[p(d),h.range()[1]]]),E=w.f().curve(w.c)([[p(l),h.range()[0]],[p(l),h.range()[1]]
[…]

### 2020-05-13 precache-manifest.eb7e28d0986a75edfc64e57ba7c41201.js
Updates

@@ -1,19 +1,19 @@
 self.__precacheManifest = (self.__precacheManifest || []).concat([
   {
-    "revision": "e8e9f154ac067d44cd81b38b8311cf09",
+    "revision": "d4a5255174b95457f1676b4fb1ba99b3",
     "url": "/pages/6894-sp20/FP-Program-Execution-Visualization/index.html"
   },
   {
-    "revision": "4f2b25ca732896f8f1fe",
+    "revision": "0fc7dcfb6af5739caf13",
     "url": "/pages/6894-sp20/FP-Program-Execution-Visualization/static/css/2.8dd801da.chunk.css"
   },
   {
-    "revision": "4f2b25ca732896f8f1fe",
-    "url": "/pages/6894-sp20/FP-Program-Execution-Visualization/static/js/2.052aba8e.chunk.js"
+    "revision": "0fc7dcfb6af5739caf13",
+    "url": "/pages/6894-sp20/FP-Program-Execution-Visualization/static/js/2.e4d82d3e.chunk.js"
   },
   {
-    "revision": "bdccee12cab5f02dc7d3",
-    "url": "/pages/6894-sp20/FP-Program-Execution-Visualization/static/js/main.47d86621.chunk.js"
+    "revision": "c19bac54104a89d6f44b",
+    "url": "/pages/6894-sp20/FP-Program-Execution-Visualization/static/js/main.a3bd9b7e.chunk.js"
   },
   {
     "revision": "1c8fb0e2dade0b210746",

### 2020-05-13 service-worker.js

@@ -14,7 +14,7 @@
 importScripts("https://storage.googleapis.com/workbox-cdn/releases/4.3.0/workbox-sw.js");
 
 importScripts(
-  "/pages/6894-sp20/FP-Program-Execution-Visualization/precache-manifest.1c9237cf565ee6320148032c0953ccc0.js"
+  "/pages/6894-sp20/FP-Program-Execution-Visualization/precache-manifest.eb7e28d0986a75edfc64e57ba7c41201.js"
 );
 
 self.addEventListener('message', (event) => {

### 2020-05-13 static/js/main.47d86621.chunk.js

@@ -1,2 +1,0 @@
-(window.webpackJsonp=window.webpackJsonp||[]).push([[0],{287:function(e,t,n){e.exports=n(542)},542:function(e,t,n){"use strict";n.r(t);var a={};n.r(a),n.d(a,"addTodo",function(){return he}),n.d(a,"deleteTodo",function(){return ve}),n.d(a,"editTodo",function(){return ge}),n.d(a,"completeTodo",function(){return Ee}),n.d(a,"completeAllTodos",function(){return Se}),n.d(a,"clearCompleted",function(){return be}),n.d(a,"setVisibilityFilter",function(){return Ie});var r=n(0),o=n.n(r),i=n(71),c=n(9),l=n(7),u=n(15),d=n(223),s=n.n(d),m=n(224),f=n.n(m),p=n(225),h=n.n(p),v=n(38),g=n(39),E=n(44),S=n(40),b=n(45),I=n(10),O=n(1),y=n.n(O),T=n(5),j=n(43),w=n(2),x=function(e){var t=e.path,n=e.stroke,a=e.fill,r=e.strokeWidth,i=e.strokeDasharray;return o.a.createElement("path",{d:t,fill:a,stroke:n,strokeWidth:r,strokeDasharray:i,opacity:.8})};x.defaultProps={stroke:"blue",fill:"none",strokeWidth:2};var k=x;function C(){var e=Object(I.a)(["\n  border: solid thin #eee;\n  cursor: pointer;\n  overflow: visible;\n"]);return C=function(){return e},e}var _=T.a.svg(C()),D=function(e){var t,n,a=e.data,r=e.width,i=e.height,c=e.setSelectedStateId,l=e.selectedStateId,u=e.jumpToState,d=e.currentStateId,s=e.resetToSelectedState,m=a.slice(-1)[0],f=a.concat([{stateId:m.stateId+1,value:m.value}]),p=w.h().domain(w.e(f,function(e){return e.stateId})).range([1,r-1]),h=w.h().domain(w.e(f,function(e){return e.value})).range([i-1,1]),v=w.f().defined(function(e){return!isNaN(e.value)}).curve(w.d).x(function(e){return p(e.stateId)}).y(function(e){return h(e.value)}),g=w.f().curve(w.c)([[p(d),h.range()[0]],[p(d),h.range()[1]]]),E=w.f().curve(w.c)([[p(l),h.range()[0]],[p(l),h.range()[1]]
[…]

### 2020-05-13 static/js/main.a3bd9b7e.chunk.js

@@ -1,0 +1,2 @@
+(window.webpackJsonp=window.webpackJsonp||[]).push([[0],{290:function(e,t,n){e.exports=n(543)},543:function(e,t,n){"use strict";n.r(t);var a={};n.r(a),n.d(a,"addTodo",function(){return Se}),n.d(a,"deleteTodo",function(){return Ie}),n.d(a,"editTodo",function(){return be}),n.d(a,"completeTodo",function(){return ye}),n.d(a,"completeAllTodos",function(){return Oe}),n.d(a,"clearCompleted",function(){return Te}),n.d(a,"setVisibilityFilter",function(){return je});var r=n(0),o=n.n(r),c=n(71),i=n(9),l=n(7),u=n(15),d=n(225),s=n.n(d),f=n(226),m=n.n(f),p=n(227),h=n.n(p),v=n(39),g=n(40),E=n(45),S=n(41),I=n(46),b=n(10),y=n(1),O=n.n(y),T=n(5),j=n(17),w=n(2),k=function(e){var t=e.path,n=e.stroke,a=e.fill,r=e.strokeWidth,c=e.strokeDasharray;return o.a.createElement("path",{d:t,fill:a,stroke:n,strokeWidth:r,strokeDasharray:c,opacity:.8})};k.defaultProps={stroke:"blue",fill:"none",strokeWidth:2};var x=k;function C(){var e=Object(b.a)(["\n  border: solid thin #eee;\n  cursor: pointer;\n  overflow: visible;\n"]);return C=function(){return e},e}var D=T.a.svg(C()),_=function(e){var t,n,a=e.data,r=e.width,c=e.height,i=e.setSelectedStateId,l=e.selectedStateId,u=e.jumpToState,d=e.currentStateId,s=e.resetToSelectedState,f=a.slice(-1)[0],m=a.concat([{stateId:f.stateId+1,value:f.value}]),p=w.h().domain(w.e(m,function(e){return e.stateId})).range([1,r-1]),h=w.h().domain(w.e(m,function(e){return e.value})).range([c-1,1]),v=w.f().defined(function(e){return!isNaN(e.value)}).curve(w.d).x(function(e){return p(e.stateId)}).y(function(e){return h(e.value)}),g=w.f().curve(w.c)([[p(d),h.range()[0]],[p(d),h.range()[1]]]),E=w.f().curve(w.c)([[p(l),h.range()[0]],[p(l),h.range()[1]]]),S=
[…]

### 2020-05-13 paper/paper.tex
little tweak to paper
finished paper

@@
 % so you remain anonymous.
-\def\plaintitle{Abstract Program Visualization for Model-View-Update
-User Interfaces}
+\def\plaintitle{Runtime Visualization for Model-View-Update GUIs}
 \def\plainauthor{Geoffrey Litt}
@@
   are most helpful, but this introduces new challenges: how does the
-  programmer specify what to visualize, and how do we visualize complex
-  data structures which aren't just primitive values? In this work, I
-  present an approach to visualizing the behavior of user interfaces
-  built with the Model-View-Update pattern. I present a prototype
-  runtime visualization system built on the Redux library and argue
-  that, by exploiting the natural abstraction characteristics of this
-  application architecture, we can create useful runtime visualizations
-  with minimal programmer effort.
+  programmer specify what to visualize, and how do they visualize
+  complex data structures which aren't just primitive values?
+
+  In this work, I present an approach to visualizing the behavior of
+  user interfaces built with the Model-View-Update pattern. I present a
+  prototype runtime visualization system built on the Redux library and
+  argue that, by exploiting the natural abstraction characteristics of
+  this application architecture, we can create useful runtime
+  visualizations with minimal programmer effort.
 \end{abstract}
@@
 focuses on low-level details: showing the values of individual
-variables, connected to individual lines of source code. This makes
-sense for small programs, and for helping novices understand the basics
-of programming languages. But these visualizations don't address the
-needs of more experienced programmers working with larger programs.
-Gaining a general understanding of a large program requires zooming out
-far above the source code level.
+variables, connected to individual lines of source code. This works well
+for small programs, and for helping novices understand the basics of
+programming. But these visualizations don't address the needs of more
+experienced programmers working with larger programs. Gaining a general
+understanding of a large program requires zooming out from individual
+lines of code.
 
@@
 creating abstract, program-specific views of runtime state or static
-code, to help someone understand the program. This idea has been
-explored in the context of teaching how algorithms work
-\autocite{brown1984,stasko1990} and in understanding the behavior of
-multithreaded Java programs \autocite{reiss2003,reiss2005}. In this
-context, a tricky challenge emerges \autocite{reiss2007}: how can we
-simultaneously tailor the visualization to the specific domain, while
-enabling the programmer to create the visualization with minimal effort?
-On the one hand, overly generic visualizations (as used in low-level
-visualization systems like Python Tutor \autocite{guo2013}) will often
-fail to capture the higher-level context of the specific program. On the
-other hand, if a visualization is too specific and takes too much work
-to create, it won't be reasonably possible for programmers to create the
-visualization.
+code, to help someone debug or understand the program. This idea has
+been explored in the context of teaching algorithms
+\autocite{brown1984,stasko1990} and understanding the behavior of
+multithreaded Java programs \autocite{reiss2003,reiss2005}. But abstract
+visualizations create a new challenge \autocite{reiss2007}: how can we
+enable the programmer to create program-specific abstract visualizations
+with minimal effort? On the one hand, overly generic visualizations (as
+used in most low-level visualization systems) will often fail to capture
+the higher-level meaning of the specific program. On the other hand, if
+a visualization takes too much work to create, it won't be realistic for
+programmers to create the visualization in practice.
 
@@
 runtime visualization systems coupled to a particular domain-specific
-framework or language. Frameworks and DSLs often impose a particular
-mental model, code architecture style, and other constraints that
-usefully narrow the space of possible programs relative to a
-general-purpose language. On the other hand, there are still many
+framework or DSL. Frameworks and DSLs occupy an intermediate place
+between general-purpose languages and specific programs. They often
+impose a particular mental model, code architecture style, and other
+constraints that usefully narrow the space of possible programs relative
+to a general-purpose language. On the other hand, there are still many
 different programs that can be built in one framework, so the effort of
@@
 
-One recent example of this strategy is visualizing runtime state in Vega
-\autocite{hoffswell2018a}, a data visualization language. Because the
-language is so domain-specific, the runtime visualization system is able
-to make many contextual assumptions leading to a useful visualization.
-
-In this work, I propose a runtime program visualization system for a
-slightly more general category of applications: user interfaces built
-with the Model-View-Update (MVU) architecture \autocite{fowler2020},
-also commonly known as the Elm Architecture \autocite{czaplicki}. MVU
+To concretely test this strategy, in this work I propose a runtime
+program visualization system for user interfaces built with the
+Model-View-Update (MVU) architecture \autocite{fowler2020}, also
+commonly known as the Elm Architecture \autocite{czaplicki}. MVU
 encourages the state of the interface to be centralized in a single data
@@
 
-This architecture has been found to have many practical benefits for
-program understanding and developer experience (e.g., making
-``time-travel debugging'' relatively trivial to implement), but I think
-it also has useful characteristics for abstract program visualization as
-well. In particular, MVU naturally encourages programmers to define
-abstractions that represent the essence of their application: 1) a
-stream of semantically meaningful events, representing important changes
-in the application, 2) a state object that represents all the core state
-of the UI, organized in a hierarchical way. My hypothesis is that it's
-possible to build a system that can visualize MVU interfaces with
-relatively little additional effort from the programmer, because of
-these natural characteristics of the architecture.
+This architecture has many practical benefits for program understanding
+and developer experience (e.g., automatically achieving time-travel
+debugging), and I think it has useful characteristics for abstract
+program visualization as well. In particular, MVU naturally encourages
+programmers to define abstractions that represent the essence of their
+application's behavior: 1) a stream of semantically meaningful events,
+2) a state object that represents all the core state of the UI. My
+hypothesis is that it is possible to visualize MVU interfaces with
+relatively little additional effort from the programmer, because the
+architecture has already required them to do much of the work of
+abstracting.
 
-To begin to explore this idea, I've prototyped a runtime visualization
-system on top of the popular Redux \autocite{zotero-621} Javascript
-library. Within the limited scope of this project, I've focused
-specifically on making a prototype specifically designed to visualize
-the state of the TodoMVC demo application. I've designed some
-visualization formats tailored to the state of that application, and
-through my own usage I've begun to gain a preliminary understanding what
-kinds of visualizations might be useful to programmers navigating the
-state of MVU applications over time.
+I've prototyped a runtime visualization system on top of the popular
+Redux \autocite{zotero-621} library, which implements MVU in Javascript.
+Within the limited scope of this project, I've focused on making a
+prototype specifically designed to visualize the state of the TodoMVC
+demo application. I've designed some visualizations tailored to the
+state of that application, and through my own usage I've begun to gain a
+preliminary understanding what kinds of visualizations might be useful
+to programmers navigating execution traces of MVU applications.
 
-Much future work remains to fully flesh out this idea, including:
-
-\begin{itemize}
-\tightlist
-\item
-  developing a crisper understanding of the needs of programmers who are
-  debugging MVU applications or trying to gain a general understanding
-  of their behavior
-\item
-  generalizing the system so that it actually works with many Redux
-  applications instead of just one, and has a way for a programmer to
-  create a program-specific visualization
-\end{itemize}
+Much future work remains to fully flesh out this idea, including
+developing a crisper understanding of the needs of programmers,
+validating this system against those needs, and generalizing the system
+so that it actually works with many Redux applications instead of just
+one demo app.
 
@@
 \autocite{victora}, Python Tutor \autocite{guo2013}, Omnicode
-\autocite{kang2017}, and Theia \autocite{pollock2019}.
+\autocite{kang2017}, Theia \autocite{pollock2019}, and Theseus
+\autocite{lieber2014}.
 
@@
 
-Other systems have explored fully abstract program visualization,
+This work aims to provide a much more abstract view of the application's
+behavior than any of these other projects, by avoiding doing any
+visualization at the level of individual lines of code.
+
+Other systems have explored this kind of abstract program visualization,
 entirely disconnected from the source code. For example, Balsa
@@
 execution of Java programs, e.g.~when different threads are running.
+
+I'm not aware of much prior research on abstract program visualization
+for user interfaces, although I still need to do a fuller literature
+review. UI performance analysis tools or debuggers like the Redux Dev
+Tools arguably fit into this category, but there aren't many tools that
+employ data visualization techniques to display the internal state of
+the application.
 
@@
 
-I had some prior experience with the Redux Dev Tools system, which
+I had some prior experience with the Redux Dev Tools debugger, which
 provides the ability to inspect application state in Redux applications.
-From this personal experience I identified two distinct use cases for a
+From this personal experience, I identified two distinct use cases for a
 runtime visualization:
@@
   trace?} This is most often helpful when debugging a particular
-  problem. Scrubbing back and forth while watching the UI change is a
-  manageable solution, but it's inefficient, and sometimes the relevant
-  state isn't directly visible in the UI.
+  problem. Scrubbing back and forth while watching the UI change is
+  often workable, but it's inefficient. Also, sometimes the relevant
+  state isn't directly visible in the UI, so I need to dig into a JSON
+  object at each point in time to understand whether I've found the
+  right point in the trace.
 \item
@@
   interested in just seeing general information about how a program is
-  behaving over time. This is helpful when explaining the system's
-  behavior to a new programmer who's preparing to work on the system, or
-  for building up my own passive understanding of a system.
+  behaving over time. For example, this is helpful when explaining the
+  system's behavior to a new programmer who's preparing to work on the
+  system, or when I'm trying to learn the basics of a codebase myself.
 \end{itemize}
@@
 the Whyline system \autocite{ko2004}), but general program understanding
-might benefit from a more passive interaction style.
+might benefit from a more passive style, more akin to reading
+documentation but accompanied by live demonstrations.
 
@@
 
-\hypertarget{sec:design}{%
-\section{Visualization Design}\label{sec:design}}
-
 \hypertarget{context-todomvc}{%
@@
 visualizations, rather than building out infrastructure, I built a
-visualization system for a specific application: the Redux
-implementation of the TodoMVC GUI benchmark. TodoMVC is a basic todo
-list UI where you can add, edit, delete, and complete todos. You can
-filter the list of todos shown to either active or completed ones.
+visualization system for a specific application: the TodoMVC GUI
+benchmark. TodoMVC is a basic todo list UI where the user can add, edit,
+delete, and complete todos. The user can also filter the list of todos
+shown to either active or completed ones.
 
-In Redux, the implementation stores an app state object as shown in
-Figure~\ref{fig:todomvc-state}. There are Redux actions corresponding to
-each of the main actions listed above, e.g.~``add todo'' and ``set
-visibility filter''. Importantly, the Redux events capture an abstract,
+The Redux implementation of TodoMVC stores an app state object which
+contains the list of todos, and the current state of the visibility
+filter. There are actions corresponding to each of the main user
+interactions listed above, e.g.~``add todo'' and ``set visibility
+filter''. Importantly, the Redux events capture an abstract,
 semantically meaningful picture of the user's interactions: when adding
@@
 Redux once the user finally adds the new todo.
-
-\begin{figure}
-\hypertarget{fig:todomvc-state}{%
-\centering
-\includegraphics[width=1.82292in,height=2.08333in]{images/todomvc-state.png}
-\caption{TodoMVC Redux application state}\label{fig:todomvc-state}
-}
-\end{figure}
 
@@
 the \emph{concrete} current state, it limits the ability of the
-programmer to create a customized abstract view. Linking the
-visualization to the Redux state is a much more abstract view than
-linking to lines of source code, but it still presents a limitation.
+programmer to create a customized abstract view, removed from the
+details of the state.
 
-As a result, in my next iteration I switched to a different layout,
-shown in Figure~\ref{fig:timeline}: a vertically stacked list of small
+In my next iteration I switched to a different layout, shown in
+Figure~\ref{fig:timeline}: a vertically stacked list of small
 visualizations of state over time. Each visualization can display an
@@
 \centering
-\includegraphics[width=3.4375in,height=2.08333in]{images/timeline.png}
+\includegraphics[width=3.4375in,height=1.77083in]{images/timeline.png}
 \caption{Timeline view of stacked visualizations}\label{fig:timeline}
@@
 which displays that data. The user can scrub through history in the
-timeline, ``pin'' the app at a particular point in time, and then use
-the separate state view to drill into the app's concrete state at that
-point.
+timeline, ``pin'' the app state at a particular point in time, and then
+use the separate state view to drill into the app's concrete state at
+that point.
 
@@
 
-Here are the specific types of visualizations I prototyped (all shown in
-Figure~\ref{fig:timeline}):
+Here I describe the specific visualizations I prototyped for the
+timeline view. These are shown in Figure~\ref{fig:timeline} from top to
+bottom. (The video demo linked on the project page might be an easier
+way to grasp the basics of each of these views)
 
 \emph{Action list}: I found that skimming a list of actions represented
-as text (``ADD\_TODO'', ``EDIT\_TODO'') required a lot of conscious
-reading effort. Instead, by choosing a colorful symbol for each action
-in the app, we can take advantage of pre-attentive processing to more
-quickly understand what actions have occurred in the execution trace. In
-this case I chose symbols for all the actions in TodoMVC; more
-generally, a programmer could specify a meaningful symbol for every
-action in their application. In some cases it might be difficult to
-choose meaningful and different symbols for all actions; falling back to
-random symbols or colored dots could work as well. In using this tool
-I've found that the symbolic action list makes it far easier to find a
-point in an execution trace that I'm looking for.
+as text (\texttt{ADD\_TODO}, \texttt{EDIT\_TODO}, etc.) required a lot
+of conscious reading effort. Instead, by choosing a colorful symbol for
+each action in the app, we can take advantage of pre-attentive
+processing to more quickly understand what actions have occurred in the
+execution trace. In this case I chose symbols for all the actions in
+TodoMVC; more generally, a programmer could specify a meaningful symbol
+for every action in their application. In some cases it might be
+difficult to choose meaningful and different symbols for all actions;
+falling back to random symbols or colored dots could work as well. In
+using this tool I've found that the symbolic action list makes it far
+easier to find a point in an execution trace that I'm looking for.
+
+\emph{Collection graph}: This visualization represents the contents of a
+collection with a series of vertically stacked dots. The size of the
+cluster of dots provides a rough sense of collection size, and the
+programmer can more carefully examine the view to get an exact count.
+
+Each dot has a color encoding for some attribute of the collection
+element; in this case I've chosen to color the dots by whether the todo
+is completed or not. Another available option is to color the dots by
+identity---each unique element gets its own color.
+
+I originally represented the list of todos with a line graph showing its
+length, but this view allows us to display an additional dimension of
+information for each todo. One corresponding weakness of this view is
+that the size encoding doesn't offer too much information for
+pre-attentive processing when there are more than a few elements so the
+relative size change is small. It's not immediately obvious where in the
+trace the number of todos changed, whereas a line graph makes it more
+obvious. (One possible improvement would be to only show the dots on a
+time step where the collection was changed.)
 
@@
 
-I implemented a working prototype version of this visualization system.
-I built a plugin on top of the existing Redux Dev Tools, which provides
-substantial infrastructure for inspecting and manipulating the state of
-a Redux application.
+I implemented a working prototype on top of the existing Redux Dev
+Tools, which provides substantial infrastructure for inspecting and
+manipulating the state of a Redux application. My tool is implemented as
+a Redux Dev Tools ``monitor'' which can plug in to those existing APIs.
 
 I used the React and Redux frameworks to implement the main skeleton of
-my system. The graphs are built in a combination of d3 and React---using
-d3 for computing scales and positions, and React for SVG rendering.
+my system. The graphs are built in a combination of d3 and React. I use
+d3 for computing scales and positions, and then React for actually
+rendering out SVGs.
 

### 2020-05-15 precache-manifest.4e6ee918701f9c2b4ca34fe9fd6311e1.js
Updates

@@ -1,0 +1,22 @@
+self.__precacheManifest = (self.__precacheManifest || []).concat([
+  {
+    "revision": "a4f4395434ae15b2880e2b65bd12729a",
+    "url": "/todomvc-vis/index.html"
+  },
+  {
+    "revision": "88aabbf0785caa0408c4",
+    "url": "/todomvc-vis/static/css/2.8dd801da.chunk.css"
+  },
+  {
+    "revision": "88aabbf0785caa0408c4",
+    "url": "/todomvc-vis/static/js/2.cc4b3b9f.chunk.js"
+  },
+  {
+    "revision": "c19bac54104a89d6f44b",
+    "url": "/todomvc-vis/static/js/main.a3bd9b7e.chunk.js"
+  },
+  {
+    "revision": "39e5ff432f5501e356c3",
+    "url": "/todomvc-vis/static/js/runtime~main.56927631.js"
+  }
+]);

### 2020-05-15 precache-manifest.eb7e28d0986a75edfc64e57ba7c41201.js

@@ -1,22 +1,0 @@
-self.__precacheManifest = (self.__precacheManifest || []).concat([
-  {
-    "revision": "d4a5255174b95457f1676b4fb1ba99b3",
-    "url": "/pages/6894-sp20/FP-Program-Execution-Visualization/index.html"
-  },
-  {
-    "revision": "0fc7dcfb6af5739caf13",
-    "url": "/pages/6894-sp20/FP-Program-Execution-Visualization/static/css/2.8dd801da.chunk.css"
-  },
-  {
-    "revision": "0fc7dcfb6af5739caf13",
-    "url": "/pages/6894-sp20/FP-Program-Execution-Visualization/static/js/2.e4d82d3e.chunk.js"
-  },
-  {
-    "revision": "c19bac54104a89d6f44b",
-    "url": "/pages/6894-sp20/FP-Program-Execution-Visualization/static/js/main.a3bd9b7e.chunk.js"
-  },
-  {
-    "revision": "1c8fb0e2dade0b210746",
-    "url": "/pages/6894-sp20/FP-Program-Execution-Visualization/static/js/runtime~main.6c763217.js"
-  }
-]);

### 2020-05-15 service-worker.js

@@ -14,7 +14,7 @@
 importScripts("https://storage.googleapis.com/workbox-cdn/releases/4.3.0/workbox-sw.js");
 
 importScripts(
-  "/pages/6894-sp20/FP-Program-Execution-Visualization/precache-manifest.eb7e28d0986a75edfc64e57ba7c41201.js"
+  "/todomvc-vis/precache-manifest.4e6ee918701f9c2b4ca34fe9fd6311e1.js"
 );
 
 self.addEventListener('message', (event) => {
@@ -33,7 +33,7 @@ workbox.core.clientsClaim();
 self.__precacheManifest = [].concat(self.__precacheManifest || []);
 workbox.precaching.precacheAndRoute(self.__precacheManifest, {});
 
-workbox.routing.registerNavigationRoute(workbox.precaching.getCacheKeyForURL("/pages/6894-sp20/FP-Program-Execution-Visualization/index.html"), {
+workbox.routing.registerNavigationRoute(workbox.precaching.getCacheKeyForURL("/todomvc-vis/index.html"), {
   
   blacklist: [/^\/_/,/\/[^\/]+\.[^\/]+$/],
 });

### 2020-05-15 static/js/runtime~main.56927631.js

@@ -1,0 +1,2 @@
+!function(e){function r(r){for(var n,i,f=r[0],l=r[1],a=r[2],p=0,s=[];p<f.length;p++)i=f[p],o[i]&&s.push(o[i][0]),o[i]=0;for(n in l)Object.prototype.hasOwnProperty.call(l,n)&&(e[n]=l[n]);for(c&&c(r);s.length;)s.shift()();return u.push.apply(u,a||[]),t()}function t(){for(var e,r=0;r<u.length;r++){for(var t=u[r],n=!0,f=1;f<t.length;f++){var l=t[f];0!==o[l]&&(n=!1)}n&&(u.splice(r--,1),e=i(i.s=t[0]))}return e}var n={},o={1:0},u=[];function i(r){if(n[r])return n[r].exports;var t=n[r]={i:r,l:!1,exports:{}};return e[r].call(t.exports,t,t.exports,i),t.l=!0,t.exports}i.m=e,i.c=n,i.d=function(e,r,t){i.o(e,r)||Object.defineProperty(e,r,{enumerable:!0,get:t})},i.r=function(e){"undefined"!==typeof Symbol&&Symbol.toStringTag&&Object.defineProperty(e,Symbol.toStringTag,{value:"Module"}),Object.defineProperty(e,"__esModule",{value:!0})},i.t=function(e,r){if(1&r&&(e=i(e)),8&r)return e;if(4&r&&"object"===typeof e&&e&&e.__esModule)return e;var t=Object.create(null);if(i.r(t),Object.defineProperty(t,"default",{enumerable:!0,value:e}),2&r&&"string"!=typeof e)for(var n in e)i.d(t,n,function(r){return e[r]}.bind(null,n));return t},i.n=function(e){var r=e&&e.__esModule?function(){return e.default}:function(){return e};return i.d(r,"a",r),r},i.o=function(e,r){return Object.prototype.hasOwnProperty.call(e,r)},i.p="/todomvc-vis/";var f=window.webpackJsonp=window.webpackJsonp||[],l=f.push.bind(f);f.push=r,f=f.slice();for(var a=0;a<f.length;a++)r(f[a]);var c=l;t()}([]);
+//# sourceMappingURL=runtime~main.56927631.js.map

### 2020-05-15 static/js/runtime~main.6c763217.js

@@ -1,2 +1,0 @@
-!function(e){function r(r){for(var n,i,a=r[0],l=r[1],f=r[2],c=0,s=[];c<a.length;c++)i=a[c],o[i]&&s.push(o[i][0]),o[i]=0;for(n in l)Object.prototype.hasOwnProperty.call(l,n)&&(e[n]=l[n]);for(p&&p(r);s.length;)s.shift()();return u.push.apply(u,f||[]),t()}function t(){for(var e,r=0;r<u.length;r++){for(var t=u[r],n=!0,a=1;a<t.length;a++){var l=t[a];0!==o[l]&&(n=!1)}n&&(u.splice(r--,1),e=i(i.s=t[0]))}return e}var n={},o={1:0},u=[];function i(r){if(n[r])return n[r].exports;var t=n[r]={i:r,l:!1,exports:{}};return e[r].call(t.exports,t,t.exports,i),t.l=!0,t.exports}i.m=e,i.c=n,i.d=function(e,r,t){i.o(e,r)||Object.defineProperty(e,r,{enumerable:!0,get:t})},i.r=function(e){"undefined"!==typeof Symbol&&Symbol.toStringTag&&Object.defineProperty(e,Symbol.toStringTag,{value:"Module"}),Object.defineProperty(e,"__esModule",{value:!0})},i.t=function(e,r){if(1&r&&(e=i(e)),8&r)return e;if(4&r&&"object"===typeof e&&e&&e.__esModule)return e;var t=Object.create(null);if(i.r(t),Object.defineProperty(t,"default",{enumerable:!0,value:e}),2&r&&"string"!=typeof e)for(var n in e)i.d(t,n,function(r){return e[r]}.bind(null,n));return t},i.n=function(e){var r=e&&e.__esModule?function(){return e.default}:function(){return e};return i.d(r,"a",r),r},i.o=function(e,r){return Object.prototype.hasOwnProperty.call(e,r)},i.p="/pages/6894-sp20/FP-Program-Execution-Visualization/";var a=window.webpackJsonp=window.webpackJsonp||[],l=a.push.bind(a);a.push=r,a=a.slice();for(var f=0;f<a.length;f++)r(a[f]);var p=l;t()}([]);
-//# sourceMappingURL=runtime~main.6c763217.js.map
